---
title: "Chapter 10"
format: html
    # theme: custom.scss
editor: visual
---

# Position scales and axes

Imports

```{r}
#| echo: false
library(tidyverse)
library(patchwork)
```

Position scales are used to control the locations of visual entities in a plot, and how those locations are mapped to data values. Every plot has two position scales, the x and y aesthetics. But in some cases, for example with `geom_histogram()` there is no need to specify y since it computes a `count` variable.

```{r}
ggplot(mpg, aes(x = displ)) + geom_histogram()
ggplot(mpg, aes(x = displ, y = after_stat(count))) + geom_histogram()

```

## 10.1 Numeric position scales

Most commonly `scale_x_continuous()` and `scale_y_continuous()`. There are others such as `scale_x_log10()` and `scale_x_reverse()`.

### 10.1.1 Limits

All scales have limit values to bound them. Useful when trying to keep views consistent across multiple plots.

```{r}
mpg_99 <- mpg %>% filter(year == 1999)
mpg_08 <- mpg %>% filter(year == 2008)

base_99 <- ggplot(mpg_99, aes(displ, hwy)) + geom_point() 
base_08 <- ggplot(mpg_08, aes(displ, hwy)) + geom_point() 

base_99 | base_08
```

Setting limits for axis for consistent scaling

```{r}
base_99 +
  scale_x_continuous(limits=c(1,7))+
  scale_y_continuous(limits=c(10,45)) |
  base_08 +
  scale_x_continuous(limits=c(1,7))+
  scale_y_continuous(limits=c(10,45))
```

Quicker way to do this is through `lims()` or `xlim()` and `ylim()`.

```{r}
base_99 + lims(x=c(1,7),y=c(10,45)) |
  base_08 + lims(x=c(1,7),y=c(10,45))
```

### 10.1.2 Zooming in

Use `coord_cartesian(xlim=c(),ylim=c())` to zoom. Using just limits will distort the plot by converting values outside of the specified range to `NA`.

```{r}
base <- ggplot(mpg,aes(drv,hwy))+
  geom_hline(yintercept = 28,color='red')+
  geom_boxplot()

base | base+coord_cartesian(ylim=c(10,35)) | base+ylim(10,35) # Distorts boxplot by converting numbers outside range to NA
```

### 10.1.3 Visual range expansion

To remove the extra bit of axes that's included through **ggplot2**, use the override with the `expand` argument and `expansion()` value.

```{r}
base <- ggplot(faithfuld,aes(waiting,eruptions))+
  geom_raster(aes(fill=density))+
  theme(legend.position='none')+
  labs(x=NULL,y=NULL)

base | base +
  scale_x_continuous(expand=expansion(0))+ # multiplied by 0 removes the expand
  scale_y_continuous(expand=expansion(0))
```

`expansion()` has both an "additive" factor, which specifies a constant space to add, and a "multiplicative" factor, which adds space defined as a proportion of the total size of the axis limit.

```{r}
# Additive
p1 <- base +
  scale_x_continuous(expand = expansion(add=3))+ # 3 units of this axis
  scale_y_continuous(expand = expansion(add=3))
# Multiplicative 
p2 <- base +
  scale_x_continuous(expand = expansion(mult = 0.2))+ # 20 % of axis range
  scale_y_continuous(expand = expansion(mult = 0.2))
# Multiplicative (5% lower end of each axes 20% upper end)
p3 <- base +
  scale_x_continuous(expand = expansion(mult = c(0.05,0.2)))+
  scale_y_continuous(expand = c(0.05, 0, 0.2, 0)) # Set directly 
p1 | p2 | p3

```

### 10.1.4 Breaks

Setting axis ticks and legend ticks are controlled with the `breaks` argument in a `scale_()` function.

```{r}
toy <- tibble(
  const = 1,
  up = 1:4,
  txt = letters[1:4],
  big = (1:4)*1000,
  log = c(2,4,10,2000)
)
toy 
```

`breaks = NULL` or a vector of data values.

```{r}
base <- ggplot(toy, aes(big,const))+
  geom_point() +
  labs(x=NULL,y=NULL)+
  scale_y_continuous(breaks=NULL)

base | base + scale_x_continuous(breaks = c(1000,2000,4000)) | base + scale_x_continuous(breaks = c(1000,1500,2000,4000))
  
```

The `breaks` argument also allows custom functions instead of vectors. Or just use the **scales** package. - `scales::breaks_extended()` creates automatic breaks for numeric axes. - `scales::breaks_log()` creates breaks appropriate for log axes. - `scales::breaks_pretty()` creates "pretty" breaks for dates/times. - `scales::breaks_width()` creates equally spaced breaks.

```{r}
base | base+scale_x_continuous(breaks=scales::breaks_extended()) | base+scale_x_continuous(breaks=scales::breaks_extended(n=2))
```

Another approach is to specify a fixed `width` that defines the spacing between breaks (`breaks_width()`).

```{r}
base +
  scale_x_continuous(breaks=scales::breaks_width(800)) |
  base + scale_x_continuous(breaks=scales::breaks_width(800,offset=200)) | # Shift all breaks 200 (left?)
  base + scale_x_continuous(breaks=scales::breaks_width(800,offset=-200)) # Shifts all breaks 200 (right?)
```

### 10.1.5 Minor breaks

Minor breaks are the unlabeled faint grid lines that appear between the major grid lines. The `minor_breaks` argument controls this.

```{r}
mb <- unique(as.numeric(1:10 %o% 10^(0:3))) # %o% generates a multiplication table
mb
```

```{r}
base <- ggplot(toy,aes(log,const)) +
  geom_point() +
  labs(x=NULL,y=NULL) +
  scale_y_continuous(breaks = NULL)

base + scale_x_log10() | base + scale_x_log10(minor_breaks = mb) # Second one helps to show logarithmic x-axis
```

### 10.1.6 Labels

`labels` argument within the `scale_()` function changes the associated label at a break.

```{r}
base <- ggplot(toy,aes(big,const))+
  geom_point()+
  labs(x=NULL,y=NULL)+
  scale_y_continuous(breaks=NULL)

base | base+scale_x_continuous(breaks=c(2000,4000),labels=c('2k','4k'))
```

Instead of using a manual `label`, a labeling function may be provided. Or use the **scales** package with some built-in function. - `scales::label_bytes()` formats numbers as kilobytes, megabytes, etc. - `scales::label_comma()` formats numbers as decimals with commas added. - `scales::label_dollar()` formats numbers as currency. - `scales::label_ordinal()` formats numbers in rank order: 1st, 2nd, 3rd etc. - `scales::label_percent()` format numbers as percentages. - `scales::label_pvalue()` formats numbers as p-values: \<0.05, \<0.1, \<0.34, etc.

```{r}
base <- ggplot(toy, aes(big, const)) + 
  geom_point() + 
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = NULL)

base |
base + scale_y_continuous(labels = scales::label_percent()) |
base + scale_y_continuous(
  labels = scales::label_dollar(prefix = "", suffix = "€")
)
```

By suppressing labels with `labels = NULL`, this will remove the labels from the axis or legend while leaving the other properties unchanged. While `breaks = NULL` is different.

```{r}
base + scale_y_continuous(breaks= NULL) | base + scale_y_continuous(labels=NULL)
```

### Exercises

1.  Recreate the following graphic:

```{r, echo = FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() + 
  scale_x_continuous("Displacement", labels = scales::unit_format(suffix = "L")) + 
  scale_y_continuous(quote(paste("Highway ", (frac(miles, gallon)))))  
```

```{r}
ggplot(mpg, aes(displ,hwy)) +
  geom_point() +
  scale_x_continuous('Displacement',labels=scales::unit_format(suffix = 'L'))+
  scale_y_continuous(quote(paste('Highway ',bgroup('(',frac(miles,gallon),')'))))

```

2.  List the three different types of object you can supply to the `breaks` argument. How do `breaks` and `labels` differ?

    1.  `NULL`

    2.  `c()`

    3.  `function{}` but not usually custom instead you use from **scales** package (e.g. `scales::breaks_pretty()`) or within **ggplot2** package.

3.  What label function allows you to create mathematical expressions? `expression()`

4.  What label function converts 1 to 1st, 2 to 2nd, and so on? `label_ordinal()`

### 10.1.7 Transformations

With continuous data, the default mapping is linear, but different transformations might be needed depending on the data.

```{r}
base <- ggplot(mpg,aes(displ, hwy)) + geom_point()

base | base+scale_x_reverse() | base+scale_y_reverse()

```

Every continuous scale takes a `trans` argument, allowing the use of a variety of transformations:

```{r}
# Converting fuel economy to fuel consumption
ggplot(mpg,aes(displ,hwy)) +
  geom_point() +
  scale_y_continuous(trans = 'reciprocal') # gallons per mile now

# Log transform x and y axes
ggplot(diamonds, aes(price,carat)) +
  geom_bin2d() +
  scale_x_continuous(trans = 'log10') +
  scale_y_continuous(trans = 'log10')

```

The transformation is carried out by a "transformer", which describes the transformation, its inverse, and how to draw the labels. `scales::trans_new()` allows construction of new transformers.

The following table lists some of the more common variants:

| Name           | Transformer                  | Function $f(x)$         | Inverse $f^{-1}(x)$  |
|------------------|-------------------|------------------|------------------|
| `"asn"`        | `scales::asn_trans()`        | $\tanh^{-1}(x)$         | $\tanh(y)$           |
| `"exp"`        | `scales::exp_trans()`        | $e ^ x$                 | $\log(y)$            |
| `"identity"`   | `scales::identity_trans()`   | $x$                     | $y$                  |
| `"log"`        | `scales::log_trans()`        | $\log(x)$               | $e ^ y$              |
| `"log10"`      | `scales::log10_trans()`      | $\log_{10}(x)$          | $10 ^ y$             |
| `"log2"`       | `scales::log2_trans()`       | $\log_2(x)$             | $2 ^ y$              |
| `"logit"`      | `scales::logit_trans()`      | $\log(\frac{x}{1 - x})$ | $\frac{1}{1 + e(y)}$ |
| `"probit"`     | `scales::probit_trans()`     | $\Phi(x)$               | $\Phi^{-1}(y)$       |
| `"reciprocal"` | `scales::reciprocal_trans()` | $x^{-1}$                | $y^{-1}$             |
| `"reverse"`    | `scales::reverse_trans()`    | $-x$                    | $-y$                 |
| `"sqrt"`       | `scales::scale_x_sqrt()`     | $x^{1/2}$               | $y ^ 2$              |

The following are equivalent

```{r}
#| fig-show: hide
#| layout-ncol: 2
#| fig-width: 4
ggplot(mpg,aes(displ,hwy))+
  geom_point() +
  scale_y_continuous(trans='reciprocal')

ggplot(mpg,aes(displ,hwy)) +
  geom_point() +
  scale_y_continuous(scales::reciprocal_trans())
```

In addition to the following, less complex, equivalent plots

```{r}
#| fig-show: hide
#| layout-ncol: 2
#| fig-width: 4
ggplot(diamonds,aes(price,carat))+
  geom_bin2d() +
  scale_x_continuous(trans = 'log10') +
  scale_y_continuous(trans = 'log10')

ggplot(diamonds, aes(price,carat)) +
  geom_bin2d() +
  scale_x_log10() +
  scale_y_log10()
```

Instead of using plots, the data can also just be modified manually.

```{r}
#| fig-show: hide
#| layout-ncol: 2
#| fig-width: 4
ggplot(mpg,aes(log10(displ),hwy)) +
  geom_point()

ggplot(mpg,aes(displ,hwy)) +
  geom_point() +
  scale_x_log10()
```

:::{.callout-note}

#### When does the transformation occur?

The transformation occurs before any statistical summaries. To transform *after* statistical computation use `coord_trans()`.

## 10.2 Date-time position scales

Special cases of numeric datatype for data/time types are `Date` (dates), `PSIXct` (for date-times), and `hms` (for time of day) classes. Most date manipulation is through the **lubridate** package.

### 10.2.1 Breaks

`date_breaks` argument allows position breaks by date units (years, months, weeks, days, hours, minutes, and seconds). For example `date_breaks = '2 weeks'` or `date_breaks = '15 years'`.

```{r}
#| layout-ncol: 2
date_base <- ggplot(economics, aes(date,psavert)) +
  geom_line(na.rm = TRUE) +
  labs(x=NULL,y=NULL)

date_base 
date_base + scale_x_date(date_breaks = '15 years')
```

:::{.callout-note}

#### Should I just use the **scales** package?

The usage of the shorthand `date_breaks = '15 years'` is a shortcut from using `breaks = scales::breaks_width('15 years')`. Use the longer form if an `offset` is desired.

```{r}
the_year <- as.Date(c('2021-01-01','2021-12-31'))
set_breaks <- scales::breaks_width('1 month')
set_breaks(the_year) # Create breaks one month apart in the range. 
```

To add in an offset for something such as a monthly salary being procssed on the eleventh day of each month.

```{r}
set_breaks <- scales::breaks_width('1 month',offset = 8)
set_breaks(the_year)
```

### 10.2.2 Minor breaks

`date_minor_breaks` argument allows specification of the minor breaks in using date units, the same way as `date_breaks` does for major breaks.

```{r}
#| layout-ncol: 2
df <- tibble(y = as_date(c('2022-01-01','2022-04-01')))
base  <- ggplot(df, aes(y=y)) +
  labs(y = NULL) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color='black'),
    panel.grid.minor = element_line(color = 'grey50')
  )

base + scale_y_date(date_breaks = '1 month')
base + scale_y_date(date_breaks = '1 month', date_minor_breaks = '1 week')
```

### 10.2.3 Labels

`labels` argument behaves similarly to the corresponding argument for numeric scales, but is often more convenient to use the `date_labels` argument.

| String | Meaning                            |
|:-------|:-----------------------------------|
| `%S`   | second (00-59)                     |
| `%M`   | minute (00-59)                     |
| `%l`   | hour, in 12-hour clock (1-12)      |
| `%I`   | hour, in 12-hour clock (01-12)     |
| `%p`   | am/pm                              |
| `%H`   | hour, in 24-hour clock (00-23)     |
| `%a`   | day of week, abbreviated (Mon-Sun) |
| `%A`   | day of week, full (Monday-Sunday)  |
| `%e`   | day of month (1-31)                |
| `%d`   | day of month (01-31)               |
| `%m`   | month, numeric (01-12)             |
| `%b`   | month, abbreviated (Jan-Dec)       |
| `%B`   | month, full (January-December)     |
| `%y`   | year, without century (00-99)      |
| `%Y`   | year, with century (0000-9999)     |

```{r}
#| layout-ncol: 2
base <- ggplot(economics, aes(date,psavert)) +
  geom_line(na.rm = TRUE) +
  labs(x = NULL, y = NULL)

base + scale_x_date(date_breaks = '5 years')
base + scale_x_date(date_breaks = '5 years', date_labels = '%y')
```

```{r}
#| layout-ncol: 3
lim <- as_date(c('2004-01-01','2005-01-01'))

base + scale_x_date(limits = lim, date_labels = '%b %y')
base + scale_x_date(limits = lim, date_labels = '%B\n%Y') # \n for line breaks
base + scale_x_date(limits = lim, labels = scales::label_date_short()) # Nice way to get Shorter named months
```

## 10.3 Discrete position scales

The following are equivalent

```{r}
#| layout-ncol: 2
ggplot(mpg, aes(x=hwy,y=class)) +
  geom_point()+
  annotate('text',x=5,y=1:7,label=1:7)

ggplot(mpg,aes(x=hwy,y=class)) +
  geom_point() +
  scale_x_continuous() +
  scale_y_discrete() +
  annotate('text',x=5,y=1:7,label=1:7)
```

Using jitter is nice feature to notice the overlapping cases which might not appear in the plot.

```{r}
ggplot(mpg,aes(x=hwy,y=class))+
  geom_jitter(width = 0, height = 0.25) + # Only vary on the categorical dimension
annotate('text',x=5,y=1:7,label=1:7)
```

### 10.3.1 Limits, breaks, and labels

`limits`, behave mostly like numeric scales, but the limits of a discrete scale are not defined in terms of endpoints and instead correspond to the set of allowable values (as characters) for that variable.

```{r}
#| layout-ncol: 3
base <- ggplot(toy, aes(const,txt))+
  geom_label(aes(label=txt)) +
  scale_x_continuous(breaks = NULL) +
  labs(x=NULL,y=NULL)

base
base + scale_y_discrete(limits =c('a','b','c','d','e')) # Note that e doesn't exist in the data but we made an entry for it on the y axis
base + scale_y_discrete(limits = c('a','d','c','b')) # The order is changed but reflected in the axis
```

`breaks`, is also mostly unchanged, but the `labels` argument allows the option of using a named vector to set the labels associated with particular values.

```{r}
#| layout-ncol: 2
base + scale_y_discrete(breaks = c('b','c')) # Only show two breaks
base + scale_y_discrete(labels = c(c='carrot',b='banana')) # Can change the label for specific values
```

:::{.callout-note}

Using `scales::label_wrap()` is nice when the categorical data contains long strings to wrap around multiple lines.

### 10.3.2 Label positions
Notice the mess on the x-axis...
```{r}
base <- ggplot(mpg,aes(manufacturer,hwy))+
  geom_boxplot()
base
```
Issue can be avoided by using `guides()` function, but in particular `guide_axis()` position aesthetic. 
```{r}
#| layout-ncol: 2
base + guides(x = guide_axis(n.dodge=3))
base + guides(x = guide_axis(angle=90))
# base + scale_x_discrete(guide = guide_axis(n.dodge = 3)) # Equivalent as above
# base + scale_x_discrete(guide = guide_axis(angle = 90)) # Equivalent as above
```
## 10.4 Binned position scales
Binned scales are when a continuous variable is sliced into multiple bins and the discretized variable is plotted. For example using `geom_bar()` and a binned position scale to approximate a histogram produced using `geom_histogram()`.
```{r}
#| layout-ncol: 2
ggplot(mpg,aes(hwy)) + geom_histogram(bins=8)
ggplot(mpg,aes(hwy)) +
  geom_bar()  +
  scale_x_binned()
```
It's more powerful when using a `geom_count()` layer, which is scaled to the number of observations in each location, but with continuous variables the overlapping does not always create nice binning behavior when values differ slightly, so by binning this is addressed. 
```{r}
#| layout-ncol: 2
ggplot(mpg,aes(displ,hwy))+
  geom_count()

ggplot(mpg,aes(displ,hwy))+
  geom_count()+
  scale_x_binned(n.breaks = 15) +
  scale_y_binned(n.breaks = 15)
```
## 10.5 Chapter Example
Attempt is to demonstrate different uses of scales and axes for data that does not translate nicely to default **ggplot2** figures. The example will be using data from the **COVID19** package (https://covid19datahub.io/) which collected a daily summary of COVID-19 cases, deaths, recovered, tests, vaccinations, and hospitalizations for 230+ countries, 760+ regions, and 12000+ administrative divisions. 
```{r}
USA_daily <- COVID19::covid19('USA') 
# Quick peak to see when to cutoff
USA_daily |> 
  ggplot(aes(confirmed)) +
  geom_histogram() +
  facet_wrap(~year(date))

covid_dates <- as.Date(c('2020-01-01','2023-06-01'))
set_breaks <- scales::breaks_width('6 months')

USA_daily |> 
  select(date,confirmed,deaths,recovered,stay_home_restrictions) |> 
  ggplot(aes(x=date,y=confirmed))+
  geom_line(na.rm = TRUE) +
  scale_x_date(breaks = set_breaks(covid_dates),date_labels = '%b-%Y',limits = as.Date(c('2020-01-01','2023-04-01')))+
  guides(x=guide_axis(angle=90))+
  labs(x='Date',y='Approximate Rate of Confirmed Cases',title = 'COVID-19 In the USA') +
  scale_y_log10(labels=scales::label_number_si())

```

