---
title: "Chapter 6"
format: html
editor: visual
---

# Chapter 6 Maps

Imports

```{r}
#| echo: false
library(tidyverse)
library(gridExtra) # multiplots
library(ggtext)
library(tidyverse)
```

## 6.1 Polygon maps
`geom_polygon()` is the the most straight forward approach to draw boundaries
```{r}
mi_counties <- map_data("county", "michigan") |> 
  dplyr::select(lon=long,lat,group,id=subregion)
head(mi_counties)
```
`lat` and `long` specify the latitude and longitude of a vertex, `id` the name of the region, and `group` for contiguous areas within a region. 
```{r}
# Example of geom_point(), kinda works but not ideal
ggplot(mi_counties, aes(lon, lat)) + 
  geom_point(size = .25, show.legend = FALSE) +
  coord_quickmap()

ggplot(mi_counties, aes(lon, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap()
```
## 6.2 Simple features maps 
Typically "longitude-latitude" data is not used in real-world mapping. Instead vector data for maps is encoded using the "simple features" standard. The `sf` package in R gives two main function, `geom_sf()` and `coord_sf()`, which can be used within ggplot. 

The `ozmaps` package, which provides maps for Australian state boundaries, local government areas, electoral boundaries, and so on will be used. 
```{r}
library(ozmaps)
library(sf)

oz_states <- ozmaps::ozmap_states
oz_states
```
The most important column is `geometry`, which specifies the spatial geometry for each of the states and territories. Each element in the `geometry` column is a multipolygon object which contains data specifying the verticies of one or more polygons that demark the border of a region. Given data with this format, `geom_sf()` and `coord_sf()` will know how to draw a map without specifying any parameters or declaring aesthetics.
```{r}
ggplot(oz_states) +
  geom_sf() + # Creates the shape geometry by finding the data in the sf object
  coord_sf() # Creates the map projection
```
### 6.2.1 Layered maps
```{r}
oz_states <- ozmaps::ozmap_states %>% filter(NAME != "Other Territories")
oz_votes <- rmapshaper::ms_simplify(ozmaps::abs_ced)
?rmapshaper::ms_simplify()
ggplot() + 
  geom_sf(data = oz_states, mapping = aes(fill = NAME), show.legend = FALSE) +
  geom_sf(data = oz_votes, fill = NA) + 
  coord_sf()
```

