---
title: "Chapter 3"
format: html
editor: visual
---

# Chapter 3 ***Individual geoms***

Chapter on Individual geoms to gain familiarization. An *individual* geeom draws a distinct graphical object for each observation (row).

Imports

```{r}
#| echo: false
library(tidyverse)
library(gridExtra) # multiplots
```

## 3.1Basic plot types

The two dimensional plots covered are: - `geom_area()` which draws an **area plot**, which is a line plot filled to the y-axis. - `geom_bar(stat='identity')` makes a **bar chart**. - `geom_line()` makes a **line plot**. The `group` aesthetic determines which observations are connected. `geom_path()` is the same but the order is based on the order the observations appear in the data. - `geom_point()` is the fan-favorite **scatterplot** - `geom_polygon()` draws polygons, which are filled paths. Each vertex is defined by a separate row in the data. - `geom_rect()`, `geom_tile()`, and `geom_raster()` draw rectangles. - `geom_text()` adds text to a plot. It requires a `label` aesthetic that provides the text to display.

```{r}
df <- data.frame(
  x = c(3, 1, 5), 
  y = c(2, 4, 6), 
  label = c("a","b","c")
)
p <- ggplot(df, aes(x, y, label = label)) + 
  labs(x = NULL, y = NULL) + # Hide axis label
  theme(plot.title = element_text(size = 12)) # Shrink plot title
p + geom_point() + ggtitle("point")
p + geom_text() + ggtitle("text")
p + geom_bar(stat = "identity") + ggtitle("bar")
p + geom_tile() + ggtitle("raster")
```

```{r}
p + geom_line() + ggtitle("line")
p + geom_area() + ggtitle("area")
p + geom_path() + ggtitle("path")
p + geom_polygon() + ggtitle("polygon")
```

### 3.1.1 Exercises

1)  What geoms would you use to draw each of the following named plots?

-   Scatterplot = `geom_point()`
-   Line chart = `geom_line()`
-   Histogram = `geom_histogram()`
-   Bar chart = `geom_bar()`
-   Pie chart = `geom_bar()+ coord_polar()`\

2)  What's the difference between the `geom_path()` and `geom_polygon()`? What's the differnce between `geom_path()` and `geom_line()` `geom_polygon()` fills in the space between the lines drawn in `geom_path()`. `geom_path()` and `geom_line()` interpret the order in a different manner.
3)  What low-level geoms are used to draw `geom_smooth()`? What about `geom_boxplot()` and `geom_violin()`? `geom_smooth()` is based on `geom_path()`, `geom_area()`, and `geom_point()`. `geom_boxplot()` is based on `geom_rect()`, `geom_line()`, and `geom_point()`. `geom_violin()` is based on `geom_area()` and `geom_path()`.
