---
title: "Chapter 10"
format: html
    # theme: custom.scss
editor: visual
---

# Position scales and axes

Imports

```{r}
#| echo: false
library(tidyverse)
library(patchwork)
```

Position scales are used to control the locations of visual entities in a plot, and how those locations are mapped to data values. Every plot has two position scales, the x and y aesthetics. But in some cases, for example with `geom_histogram()` there is no need to specify y since it computes a `count` variable.

```{r}
ggplot(mpg, aes(x = displ)) + geom_histogram()
ggplot(mpg, aes(x = displ, y = after_stat(count))) + geom_histogram()

```

## 10.1 Numeric position scales

Most commonly `scale_x_continuous()` and `scale_y_continuous()`. There are others such as `scale_x_log10()` and `scale_x_reverse()`.

### 10.1.1 Limits

All scales have limit values to bound them. Useful when trying to keep views consistent across multiple plots.

```{r}
mpg_99 <- mpg %>% filter(year == 1999)
mpg_08 <- mpg %>% filter(year == 2008)

base_99 <- ggplot(mpg_99, aes(displ, hwy)) + geom_point() 
base_08 <- ggplot(mpg_08, aes(displ, hwy)) + geom_point() 

base_99 | base_08
```

Setting limits for axis for consistent scaling

```{r}
base_99 +
  scale_x_continuous(limits=c(1,7))+
  scale_y_continuous(limits=c(10,45)) |
  base_08 +
  scale_x_continuous(limits=c(1,7))+
  scale_y_continuous(limits=c(10,45))
```

Quicker way to do this is through `lims()` or `xlim()` and `ylim()`.

```{r}
base_99 + lims(x=c(1,7),y=c(10,45)) |
  base_08 + lims(x=c(1,7),y=c(10,45))
```

### 10.1.2 Zooming in

Use `coord_cartesian(xlim=c(),ylim=c())` to zoom. Using just limits will distort the plot by converting values outside of the specified range to `NA`.

```{r}
base <- ggplot(mpg,aes(drv,hwy))+
  geom_hline(yintercept = 28,color='red')+
  geom_boxplot()

base | base+coord_cartesian(ylim=c(10,35)) | base+ylim(10,35) # Distorts boxplot by converting numbers outside range to NA
```

### 10.1.3 Visual range expansion

To remove the extra bit of axes that's included through **ggplot2**, use the override with the `expand` argument and `expansion()` value.

```{r}
base <- ggplot(faithfuld,aes(waiting,eruptions))+
  geom_raster(aes(fill=density))+
  theme(legend.position='none')+
  labs(x=NULL,y=NULL)

base | base +
  scale_x_continuous(expand=expansion(0))+ # multiplied by 0 removes the expand
  scale_y_continuous(expand=expansion(0))
```

`expansion()` has both an "additive" factor, which specifies a constant space to add, and a "multiplicative" factor, which adds space defined as a proportion of the total size of the axis limit.

```{r}
# Additive
p1 <- base +
  scale_x_continuous(expand = expansion(add=3))+ # 3 units of this axis
  scale_y_continuous(expand = expansion(add=3))
# Multiplicative 
p2 <- base +
  scale_x_continuous(expand = expansion(mult = 0.2))+ # 20 % of axis range
  scale_y_continuous(expand = expansion(mult = 0.2))
# Multiplicative (5% lower end of each axes 20% upper end)
p3 <- base +
  scale_x_continuous(expand = expansion(mult = c(0.05,0.2)))+
  scale_y_continuous(expand = c(0.05, 0, 0.2, 0)) # Set directly 
p1 | p2 | p3

```

### 10.1.4 Breaks

Setting axis ticks and legend ticks are controlled with the `breaks` argument in a `scale_()` function.

```{r}
toy <- tibble(
  const = 1,
  up = 1:4,
  txt = letters[1:4],
  big = (1:4)*1000,
  log = c(2,4,10,2000)
)
toy 
```

`breaks = NULL` or a vector of data values.

```{r}
base <- ggplot(toy, aes(big,const))+
  geom_point() +
  labs(x=NULL,y=NULL)+
  scale_y_continuous(breaks=NULL)

base | base + scale_x_continuous(breaks = c(1000,2000,4000)) | base + scale_x_continuous(breaks = c(1000,1500,2000,4000))
  
```

The `breaks` argument also allows custom functions instead of vectors. Or just use the **scales** package. - `scales::breaks_extended()` creates automatic breaks for numeric axes. - `scales::breaks_log()` creates breaks appropriate for log axes. - `scales::breaks_pretty()` creates "pretty" breaks for dates/times. - `scales::breaks_width()` creates equally spaced breaks.

```{r}
base | base+scale_x_continuous(breaks=scales::breaks_extended()) | base+scale_x_continuous(breaks=scales::breaks_extended(n=2))
```

Another approach is to specify a fixed `width` that defines the spacing between breaks (`breaks_width()`).

```{r}
base +
  scale_x_continuous(breaks=scales::breaks_width(800)) |
  base + scale_x_continuous(breaks=scales::breaks_width(800,offset=200)) | # Shift all breaks 200 (left?)
  base + scale_x_continuous(breaks=scales::breaks_width(800,offset=-200)) # Shifts all breaks 200 (right?)
```

### 10.1.5 Minor breaks

Minor breaks are the unlabeled faint grid lines that appear between the major grid lines. The `minor_breaks` argument controls this.

```{r}
mb <- unique(as.numeric(1:10 %o% 10^(0:3))) # %o% generates a multiplication table
mb
```

```{r}
base <- ggplot(toy,aes(log,const)) +
  geom_point() +
  labs(x=NULL,y=NULL) +
  scale_y_continuous(breaks = NULL)

base + scale_x_log10() | base + scale_x_log10(minor_breaks = mb) # Second one helps to show logarithmic x-axis
```

### 10.1.6 Labels

`labels` argument within the `scale_()` function changes the associated label at a break.

```{r}
base <- ggplot(toy,aes(big,const))+
  geom_point()+
  labs(x=NULL,y=NULL)+
  scale_y_continuous(breaks=NULL)

base | base+scale_x_continuous(breaks=c(2000,4000),labels=c('2k','4k'))
```

Instead of using a manual `label`, a labeling function may be provided. Or use the **scales** package with some built-in function. - `scales::label_bytes()` formats numbers as kilobytes, megabytes, etc. - `scales::label_comma()` formats numbers as decimals with commas added. - `scales::label_dollar()` formats numbers as currency. - `scales::label_ordinal()` formats numbers in rank order: 1st, 2nd, 3rd etc. - `scales::label_percent()` format numbers as percentages. - `scales::label_pvalue()` formats numbers as p-values: \<0.05, \<0.1, \<0.34, etc.

```{r}
base <- ggplot(toy, aes(big, const)) + 
  geom_point() + 
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = NULL)

base |
base + scale_y_continuous(labels = scales::label_percent()) |
base + scale_y_continuous(
  labels = scales::label_dollar(prefix = "", suffix = "€")
)
```

By suppressing labels with `labels = NULL`, this will remove the labels from the axis or legend while leaving the other properties unchanged. While `breaks = NULL` is different.

```{r}
base + scale_y_continuous(breaks= NULL) | base + scale_y_continuous(labels=NULL)
```

### Exercises

1.  Recreate the following graphic:

```{r, echo = FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() + 
  scale_x_continuous("Displacement", labels = scales::unit_format(suffix = "L")) + 
  scale_y_continuous(quote(paste("Highway ", (frac(miles, gallon)))))  
```

```{r}
ggplot(mpg, aes(displ,hwy)) +
  geom_point() +
  scale_x_continuous('Displacement',labels=scales::unit_format(suffix = 'L'))+
  scale_y_continuous(quote(paste('Highway ',bgroup('(',frac(miles,gallon),')'))))

```

2.  List the three different types of object you can supply to the `breaks` argument. How do `breaks` and `labels` differ?

    1.  `NULL`

    2.  `c()`

    3.  `function{}` but not usually custom instead you use from **scales** package (e.g. `scales::breaks_pretty()`) or within **ggplot2** package.

3.  What label function allows you to create mathematical expressions? `expression()`

4.  What label function converts 1 to 1st, 2 to 2nd, and so on? `label_ordinal()`

### 10.1.7 Transformations

With continuous data, the default mapping is linear, but different transformations might be needed depending on the data.

```{r}
base <- ggplot(mpg,aes(displ, hwy)) + geom_point()

base | base+scale_x_reverse() | base+scale_y_reverse()

```

Every continuous scale takes a `trans` argument, allowing the use of a variety of transformations:

```{r}
# Converting fuel economy to fuel consumption
ggplot(mpg,aes(displ,hwy)) +
  geom_point() +
  scale_y_continuous(trans = 'reciprocal') # gallons per mile now

# Log transform x and y axes
ggplot(diamonds, aes(price,carat)) +
  geom_bin2d() +
  scale_x_continuous(trans = 'log10') +
  scale_y_continuous(trans = 'log10')

```

The transformation is carried out by a "transformer", which describes the transformation, its inverse, and how to draw the labels. `scales::trans_new()` allows construction of new transformers.

The following table lists some of the more common variants:

| Name           | Transformer                  | Function $f(x)$         | Inverse $f^{-1}(x)$  |
|------------------|--------------------|------------------|------------------|
| `"asn"`        | `scales::asn_trans()`        | $\tanh^{-1}(x)$         | $\tanh(y)$           |
| `"exp"`        | `scales::exp_trans()`        | $e ^ x$                 | $\log(y)$            |
| `"identity"`   | `scales::identity_trans()`   | $x$                     | $y$                  |
| `"log"`        | `scales::log_trans()`        | $\log(x)$               | $e ^ y$              |
| `"log10"`      | `scales::log10_trans()`      | $\log_{10}(x)$          | $10 ^ y$             |
| `"log2"`       | `scales::log2_trans()`       | $\log_2(x)$             | $2 ^ y$              |
| `"logit"`      | `scales::logit_trans()`      | $\log(\frac{x}{1 - x})$ | $\frac{1}{1 + e(y)}$ |
| `"probit"`     | `scales::probit_trans()`     | $\Phi(x)$               | $\Phi^{-1}(y)$       |
| `"reciprocal"` | `scales::reciprocal_trans()` | $x^{-1}$                | $y^{-1}$             |
| `"reverse"`    | `scales::reverse_trans()`    | $-x$                    | $-y$                 |
| `"sqrt"`       | `scales::scale_x_sqrt()`     | $x^{1/2}$               | $y ^ 2$              |

The following are equivalent

```{r}
#| fig-show: hide
#| layout-ncol: 2
#| fig-width: 4
ggplot(mpg,aes(displ,hwy))+
  geom_point() +
  scale_y_continuous(trans='reciprocal')

ggplot(mpg,aes(displ,hwy)) +
  geom_point() +
  scale_y_continuous(scales::reciprocal_trans())
```

In addition to the following, less complex, equivalent plots

```{r}
#| fig-show: hide
#| layout-ncol: 2
#| fig-width: 4
ggplot(diamonds,aes(price,carat))+
  geom_bin2d() +
  scale_x_continuous(trans = 'log10') +
  scale_y_continuous(trans = 'log10')

ggplot(diamonds, aes(price,carat)) +
  geom_bin2d() +
  scale_x_log10() +
  scale_y_log10()
```

Instead of using plots, the data can also just be modified manually.

```{r}
#| fig-show: hide
#| layout-ncol: 2
#| fig-width: 4
ggplot(mpg,aes(log10(displ),hwy)) +
  geom_point()

ggplot(mpg,aes(displ,hwy)) +
  geom_point() +
  scale_x_log10()
```

:::{.callout-note}

#### When does the transformation occur?

The transformation occurs before any statistical summaries. To transform *after* statistical computation use `coord_trans()`.


## 10.2 Date-time position scales

Special cases of numeric datatype for data/time types are `Date` (dates), `PSIXct` (for date-times), and `hms` (for time of day) classes. Most date manipulation is through the **lubridate** package.

### 10.2.1 Breaks
