---
title: "Chapter 2"
format: html
editor: visual
---

# Chapter 2

Mostly an introduction chapter.

Imports
```{r}
#| echo: false
library(tidyverse)
```


## 2.2 Fuel Economy Data
One of the bread and butter datasets we will use for visualizations
```{r}
head(mpg)
```


Initial questions: How are the engine and fuel economy related? Do certain manufacturers care more about the economy more than others? (Or other aspects of the car like power?) Has fuel economy improved in the last 10 years? Is the fuel improvements mostly from a few manufacturers or all?

### 2.2.1 Exercises
1) List five functions that you could use to get more information about the `mpg` dataset.
```{r}
# Looking at mpg across manufacturers
mpg |>
  group_by(manufacturer) |> 
  summarise(mean=mean(cty),std = sd(cty),med=median(cty)) |> 
  arrange(desc(mean))
# Lets just do a boxplot and that's five functions
mpg |>
  group_by(manufacturer) |> 
  ggplot(aes(x=manufacturer,y=cty)) +
  geom_boxplot() +
  labs(x='Manufacturer',y='City MPG')+
  theme(axis.text.x  = element_text(angle=-90, hjust=0.5, vjust=0, size=11,colour="black"))

```
2) How can you find out what other datasets are included with ggplot2?
```{r}
data(package='ggplot2')
```
3) Apart from the US, most countries use fuel consumption (fuel consumed over fixed distance) rather than fuel economy (distance traveled with fixed amount of fuel). How could you convert cty and hwy into the European standard of 1/100 km?
```{r}
mpg |> 
  mutate(cty_miles_per_liter = cty/3.785412,
         cty_kilometers_per_liter = cty_miles_per_liter*1.609344,
         cty_liters_per_100km = 100/cty_kilometers_per_liter)
```
4) Which manufacturer has the most models in this dataset? Which model has the most variations? Does your answer change if you remove the redundant specification of drive train (e.g. "pathfinder 4wd", "a4 quattro") from the model name
```{r}
mpg |> 
  group_by(manufacturer) |> 
  summarize(count = n()) |> 
  arrange(desc(count))

mpg |> 
  group_by(model) |> 
  summarize(count = n()) |> 
  arrange(desc(count))

mpg |>
  mutate(across('model',str_replace,'a4 quattro','a4'),across('model',str_remove,'4wd')) |>
  group_by(model) |>
  summarize(count = n()) |>
  arrange(desc(count))
```
## 2.3 Key Components
Key components to every ggplot figure
```{r}
ggplot(mpg,aes(x=displ,y=hwy,color=cyl)) + # Added a little color
  geom_point()
```
### 2.3.1 Exercises
1) How would you describe the relationship between cty and hwy? Do you have any concerns about drawing conclusions from that plot?
```{r}
ggplot(mpg,aes(cty,hwy))+
  geom_point()
```

2) What does `ggplot(mpg,aes(model,manufacturer)) + geom_point()` show? Is it useful? How would you modify it so it is?
```{r}
ggplot(mpg,aes(model,hwy)) + 
  geom_point() +
  facet_grid(~manufacturer) +
  theme(axis.ticks.x = element_blank(),strip.background.x = element_blank())
# I can't figure out how to get rid of the axis ticks for model at the bottom... 

```

3) Describe the data, aesthetic mappings, and layers in the following
```{r}
ggplot(mpg,aes(cty,hwy))+geom_point()
ggplot(diamonds,aes(carat,price))+geom_point()
ggplot(economics, aes(date,unemploy))+geom_line()
ggplot(mpg,aes(cty))+geom_histogram()
```



